---
title: "Challenge Data Meta Analysis"
date: "Last modified on `r Sys.Date()`"
output: 
  html_document:
    highlight: tango
    toc_float: true
    toc_collapsed: true
    theme: united
    toc: yes
    toc_depth: 3  
---

<style>
body {
  font-family: sans-serif;
  font-size: 18px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, include=FALSE}
source("load.R")
source("functions.R")
collated_url <- "https://docs.google.com/spreadsheets/d/1oiUUcNZdZ-EccKtZpG_o7DLiKM8F5n69YScqjkpC0X8"
```

## Background

---

DREAM challenges lead the field of crowdsourced benchmarking in biomedical sciences. Between 2012 - 2021, DREAM has organized and completed almost 45 challenges. While most of the challenge outcomes have been documented, evaluated, and published in well-known peer reviewed journals, it is time to review the challenges themselves to understand more about what makes challenges successful and what could be improved upon.

## Question

---

Do **DREAM** challenges form clusters according to features?

## Input Data

---

[Collated Challenge Features](https://docs.google.com/spreadsheets/d/1oiUUcNZdZ-EccKtZpG_o7DLiKM8F5n69YScqjkpC0X8/edit#gid=0)

```{r}
# need config your gmail if it is the first time running
suppressMessages(
  dream_data <- read_sheet(collated_url, range = "DREAM!A2:AJ47")
)

glimpse(dream_data)
```

### Current Selected/Used Features

```{r}
num_feats <- c("challengeAge", "numberTotalSubmissions", "numberOfTotalRegistrants", "numberOfDataTypes", 
              colnames(dream_data)[25:36]) # all user engagement data
cate_feats <- c("dataTypeTiers", "sponsorTiers", "contributorTiers")
logi_feats <- c("modelContainerization", "monetaryIncentive", "speakingIncentive")

all_feats <- c(num_feats, cate_feats, logi_feats)
metadata <- dream_data[, c("challengeName", all_feats)] %>% tibble::column_to_rownames("challengeName")
glimpse(metadata)
```

<br>

- **Numerical Data**: ``r num_feats``
- **Categorical Data**: ``r cate_feats``
- **Logical Data**: ``r logi_feats``

```{r}
num_inx <- which(colnames(metadata) %in% num_feats)
```

## Data Transformation

Most features are strong positively skewed.
```{r}
num_dat <- metadata[, num_inx] %>% 
  select(-c(numberTotalSubmissions, numberOfTotalRegistrants)) %>%
  na.omit()

plot(hist(as.matrix(num_dat), breaks = 600), main = "Histogram on Continuous Data", xlab = NULL, xlim = c(0, 500))
```

## Exploratory Heatmap {.tabset}

---

### Keep All Missing Data

```{r}
all_dat <- metadata
all_dat[, -num_inx] <- lapply(all_dat[, -num_inx], as.factor)
all_dat <- dplyr::select(all_dat, -c("numberTotalSubmissions","numberOfTotalRegistrants"))
all_gower_dist <- daisy(all_dat, metric = "gower", type = list(logratio = 2))
```

#### Check with Silhouette
```{r}
all_gower_mat <- as.matrix(all_gower_dist)
plotSilhouette(all_gower_mat)
# 
all_pam_dat <- pam(all_gower_dist, diss = TRUE, k = 2)
all_pam_clusters <- factor(all_pam_dat$clustering)
```
#### tsne
```{r}
# challenges_c1 <- names(pam_clusters[pam_clusters == 1])
# challenges_c2 <- names(pam_clusters[pam_clusters == 2])

plotDistTSNE(all_gower_dist, 
             cluster = all_pam_clusters, 
             label = strtrim(rownames(all_dat), 9),
             seed = 1234)

plotDistTSNE(all_gower_dist, 
             cluster = all_pam_clusters, 
             label = all_dat$challengeAge,
             seed = 1234)
```

#### summary
```{r}
all_quan_dat <- metadata[, num_inx]
all_quan_dat$clusters <- all_pam_clusters
all_quan_dat %>%
  group_by(clusters) %>% 
  summarise_each(funs(sum(is.na(.)))) %>%
  t() %>%
  DT::datatable()
```

#### boxplot
```{r,fig.height=8}
plotQuanBoxplot(all_quan_dat)
```

### Removed User Engagement Features
```{r}
rm_ue_dat <- metadata[, c(1,4, 17:22)]
rm_ue_dat[, 3:8] <- lapply(rm_ue_dat[, 3:8], as.factor)
rm_ue_gower_dist <- daisy(rm_ue_dat,
                    metric = "gower" ,
                    type = list(logratio = 2))
```

#### silhouette
```{r}
rm_ue_gower_mat <- as.matrix(rm_ue_gower_dist)
plotSilhouette(rm_ue_gower_mat)
```

```{r}
rm_ue_pam_dat <- pam(rm_ue_gower_dist, diss = TRUE, k = 5)
rm_ue_pam_clusters <- factor(rm_ue_pam_dat$clustering)
```

#### tsne
```{r}
plotDistTSNE(rm_ue_gower_dist, 
             cluster = rm_ue_pam_clusters, 
             label = strtrim(rownames(rm_ue_dat), 9),
             seed = 3345)
```

```{r}
rm_ue_quan_dat <- metadata[, num_inx]
rm_ue_quan_dat$clusters <- as.factor(rm_ue_pam_clusters)
```

#### boxplot
```{r, fig.height=8}
plotQuanBoxplot(rm_ue_quan_dat)
```

```{r}
rm_ue_qual_dat3 <- metadata[, c(1,4, 17:22)]
rm_ue_qual_dat3$clusters <- as.factor(rm_ue_pam_clusters)
pp <- plotQualBar(rm_ue_qual_dat3)

(pp[[4]] + pp[[5]]) + plot_layout(guides = "collect")
((pp[[1]] + pp[[2]] + pp[[3]]) + plot_layout(guides = "collect"))
pp[[6]]
```



#### PCA

```{r}
# FAMD (Factor Analysis of Mixed Data)
pca_dat <- rm_ue_dat
pca_dat[, 1:2] <- scale(as.data.frame(apply(pca_dat[, 1:2], 2, transform_log2)))
# pca_dat[, num_inx] <- scale(as.data.frame(apply(pca_dat[, num_inx], 2, transform_log2)))
# pca_dat[, cate_feats] <- lapply(pca_dat[, cate_feats], function(x) as.factor(x))
# impute missing values
pca_dat <- replace_na(pca_dat, "Missing")
# get PCA data
famd_res <- FAMD(pca_dat, graph = FALSE)
fviz_contrib(famd_res, "var", axes = 1)
# Contribution to the second dimension
fviz_contrib(famd_res, "var", axes = 2)
```

### Use Challenges from cluster 2 

```{r}
new_chlg_dat <- metadata[24:45, ]
new_chlg_dat[, -num_inx] <- lapply(new_chlg_dat[, -num_inx], as.factor)
new_chlg_dat <- dplyr::select(new_chlg_dat, -c("numberTotalSubmissions","numberOfTotalRegistrants"))
new_chlg_gower_dist <- 
  daisy(new_chlg_dat,
        metric = "gower" ,
        type = list(logratio = 2))
```

#### silhouette
```{r}
new_chlg_gower_mat <- as.matrix(new_chlg_gower_dist)
plotSilhouette(new_chlg_gower_mat, k = 15)
```

```{r}
new_chlg_pam_dat <- pam(new_chlg_gower_dist, diss = TRUE, k = 2)
new_chlg_pam_clusters <- factor(new_chlg_pam_dat$clustering)
```

#### tsne
```{r}
plotDistTSNE(new_chlg_gower_dist, 
             cluster = new_chlg_pam_clusters, 
             label = strtrim(rownames(new_chlg_dat), 9),
             seed = 1234)
```

```{r}
new_chlg_quan_dat <- metadata[24:45, num_inx]
new_chlg_quan_dat$clusters <- as.factor(new_chlg_pam_clusters)
```

#### boxplot
```{r}
plotQuanBoxplot(new_chlg_quan_dat)
```


```{r}
pheatmap(new_chlg_gower_mat, scale = "none",show_colnames = F)
```


```{r,fig.height=10}
hc <- lapply(c("single", "average", "complete"), function(clust) {
  hclust(new_chlg_gower_dist, method = clust)
})
par(mar = c(0,0,4,15))
# {
#   as.dendrogram(hc[[1]]) %>%
#     plot(axes = FALSE, horiz = TRUE, main = "single")
#   # abline(v = 0.2, lty = 2)
# }
# {
#   as.dendrogram(hc[[2]]) %>%
#     set("labels_col", k = 4) %>%
#     set("branches_k_color", k = 4) %>%
#     plot(axes = FALSE, horiz = TRUE, main = "average")
#   abline(v = 0.35, lty = 2)
# }
{
  as.dendrogram(hc[[3]]) %>%
    set("labels_col", k = 5) %>%
    set("branches_k_color", k = 5) %>%
    plot(axes = TRUE, horiz = TRUE, main = "complete")
  abline(v = 0.42, lty = 2)
}
```
#### hc-tsne
```{r}
hc_clusters <- cutree(hc[[3]], k = 5)
plotDistTSNE(new_chlg_gower_dist, 
             cluster = factor(hc_clusters), 
             label = strtrim(rownames(new_chlg_dat), 9),
             seed = 3334)
```
#### PCA
```{r}
# FAMD (Factor Analysis of Mixed Data)
# pca_dat <- new_chlg_dat
# pca_dat[, 1:14] <- scale(as.data.frame(apply(pca_dat[, 1:14], 2, transform_log2)))
# pca_dat[, 15:20] <- lapply(pca_dat[, 15:20], function(x) as.factor(x))
# # impute missing values
# res.impute <- imputeFAMD(pca_dat)
# # get PCA data
# famd_res <- FAMD(pca_dat, tab.disj = res.impute$tab.disj, graph = FALSE)
# fviz_screeplot(famd_res, addlabels = TRUE, ylim = c(0, 50)) + labs(title = "Explained Variance Pct each PC")
# fviz_contrib(famd_res, "var", axes = 1)
# # Contribution to the second dimension
# fviz_contrib(famd_res, "var", axes = 2)
# f1 <- famd_res$var$contrib[,1] %>% sort(decreasing = T) %>% head(9) %>% names
# f2 <- famd_res$var$contrib[,2] %>% sort(decreasing = T) %>% head(3) %>% names
# unique(f1,f2)
# 
# 
# pairs(metadata[, c("numberTotalSubmissions", "numberOfTotalRegistrants", unique(f1,f2))])

```

#### hc-boxplot
```{r}
new_chlg_quan_dat <- metadata[24:45, num_inx]
new_chlg_quan_dat$clusters <- as.factor(hc_clusters)
data.frame(hc_clusters %>% sort)
```

#### boxplot
```{r, fig.height=8}
plotQuanBoxplot(new_chlg_quan_dat)
```
```{r}
new_chlg_qual_dat3 <- metadata[24:45, -num_inx]
new_chlg_qual_dat3$clusters <- as.factor(hc_clusters)
new_chlg_qual_dat3 <- new_chlg_qual_dat3 %>%
  tibble::rownames_to_column("name") %>%
  replace_na("Missing") %>%
  separate_rows(dataTypeTiers) %>%
  separate_rows(sponsorTiers) %>%
  separate_rows(contributorTiers) %>%
  gather("key", "value", -c(name, clusters)) %>%
  distinct() %>%
  mutate(key = as.factor(key))

p3 <- lapply(c("dataTypeTiers", 
              "sponsorTiers", "contributorTiers", 
              "modelContainerization", "monetaryIncentive", "speakingIncentive"), function(feat) {
  
  p.data <- filter(new_chlg_qual_dat3, key == feat)
  
  if (feat %in% c("sponsorTiers", "contributorTiers")) {
    p.data$value <- factor(p.data$value, 
                     levels = c("Academic", "Government","Industrial","Foundation",
                                "Other", "Missing"))
  }
  out <- ggplot(p.data,
    aes(clusters, fill = value)) + 
    geom_bar(stat = "count") +
    labs(y = NULL) + 
    facet_wrap(~key, ncol = 4, scale="free") +
    ggsci::scale_fill_jco() +
    theme_bw()
  
  # if (feat == "dataTypeTiers") {
  #   out <- out + scale_fill_brewer(palette = "Dark2")
  # }
  return(out)
})
  (( p3[[2]] + p3[[3]]) + plot_layout(guides = "collect"))
((p3[[4]] + p3[[5]] + p3[[6]]) + plot_layout(guides = "collect"))

p3[[1]]
```

