---
title: "Challenge Data Meta Analysis"
date: "Last modified on `r Sys.Date()`"
output: 
  html_document:
    highlight: tango
    toc_float: true
    toc_collapsed: true
    theme: united
    toc: yes
    toc_depth: 3  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
update_res <- FALSE
figure_dir <- "../figures"
```

```{r, include=FALSE}
suppressPackageStartupMessages({
  library("tidyverse")
  library("data.table")
  library("splitstackshape")
  # library("caret")
  library("mltools")
  library("pheatmap")
  # library("googlesheets4")
  library("pheatmap")
  library("dendextend")
  library("patchwork")
  library("cluster")
  library("Rtsne")
  library("ggpubr")
  library("ggsci")
})
source("functions.R")
```

## Background

---

DREAM challenges lead the field of crowdsourced benchmarking in biomedical sciences. Between 2012 - 2021, DREAM has organized and completed almost 45 challenges. While most of the challenge outcomes have been documented, evaluated, and published in well-known peer reviewed journals, it is time to review the challenges themselves to understand more about what makes challenges successful and what could be improved upon.

## Question

---

- Do **DREAM** challenges form clusters according to features?

## Input Data

---

[Collated Challenge Features](https://docs.google.com/spreadsheets/d/1oiUUcNZdZ-EccKtZpG_o7DLiKM8F5n69YScqjkpC0X8/edit#gid=0)

```{r}
dream_data <- fread("../data/dream_v1.csv", data.table = FALSE, header = TRUE) %>%
  janitor::row_to_names(1) %>%
  # remove the BraTS for now
  filter(challengeName != "International Brain Tumor Segmentation")

glimpse(dream_data)
```

### Current Selected/Used Features

```{r categorize}
# categorize different features
forum_feats <- colnames(dream_data)[27:38] # all user engagement data
num_feats <- c("challengeAge",
               "numberTotalSubmissions", 
               "numberOfTotalRegistrants", 
               "numberOfDataTypes",
               "numberOfSubChallenges",
               forum_feats
               )
cate_feats <- c("dataTypeTiers", "sponsorTiers", "contributorTiers")
logi_feats <- c("modelContainerization", "monetaryIncentive", "speakingIncentive")

all_feats <- c(num_feats, cate_feats, logi_feats)

metadata <- dream_data[, c("challengeName", all_feats)] %>%
  tibble::column_to_rownames("challengeName") %>%
  # clean up types from exported csv
  mutate_at(num_feats, as.numeric) %>%
  mutate_at(c(cate_feats, logi_feats), as.factor)

glimpse(metadata)
```

<br>

- **Numerical Data**: ``r num_feats``
- **Categorical Data**: ``r cate_feats``
- **Logical Data**: ``r logi_feats``

## Data Transformation

The continuous data is strong positively skewed. We will need to apply the log transformation on continuous data before performing clustering or PCA.

```{r}
# save the index of numeric features
num_inx <- which(colnames(metadata) %in% num_feats)
num_data <- metadata[, num_inx] %>% 
  select(-c(numberTotalSubmissions, numberOfTotalRegistrants)) %>%
  na.omit() %>%
  as.matrix()

# pdf(file.path(figure_dir, "histogram_continuous_data.pdf"), width = 10, height = 4)

par(mfrow = c(1, 2), mgp = c(2.4, 0.6, 0))
hist(num_data, breaks = 100, main = "positive skewed", 
     xlab = NULL, xlim = c(0, 500), col = "lightblue", 
     cex.main = 1.2, cex.lab = 1.2, cex.axis = 1)

# log2 tranformation
num_data_log2 <- log2(num_data + 0.1)
hist(num_data_log2, breaks = 20, main = "after log transformation", 
     xlab = NULL, ylab = NULL,  xlim = c(0, 11), col = rgb(1, 0, 0, 0.5), 
     cex.main = 1.2, cex.lab = 1.2, cex.axis = 1)

# dev.off()
```

## Exploratory Heatmap {.tabset}

---

I have initially one-hot encode categorical features and did clustering with the euclidean distance. However, various types of categorical data introduces large sparsity in the heatmap and one-hot encoding is not easy to interpret with many missing data.

Alternatively, we chose "Gower Distance". Gower is a distance method popularly used for mixed type data:
- It can handle missing data
- It works with mixed data types. It calculates similarity coefficients using specific metrics for each data type (continuous, ordinal, categorical)
- The distance matrix it creates is easily interpretable on a heatmap


### All challenges

#### Gower distance heatmap
Create heatmap using gower distance to explore pattern. The number of participants and registrants were excluded. The dissimilarity matrix/distance matrix is calculated using `cluster::daisy` function. 

```{r}
# make a copy of metadata
all_dat <- metadata
# all_dat$challengeAge <- as.factor(all_dat$challengeAge)

# calculate gower distance with all features
# apply log2 and gower distance
all_gower_dist <- daisy(all_dat, metric = "gower", type = list(logratio = 2))
all_gower_mat <- as.matrix(all_gower_dist)


heatmap_all_gower <- pheatmap(all_gower_mat, show_colnames = FALSE, show_rownames = FALSE, cutree_cols = 3)

if (update_res) {
  heatmap_all_gower <- pheatmap(all_gower_mat, show_colnames = FALSE, cutree_cols = 3, fontsize = 24) # poster
  ggsave(file.path(figure_dir, "heatmap_all_gower.pdf"), heatmap_all_gower, width = 24, height = 16 , limitsize = FALSE)
}
```

#### Partition Around Medoids (PAM)

We chose Partition Around Medoids (PAM) to further determine the number of clusters. According to the Silhouette plot, when k = 2, challenges will have the highest Silhouette width.

```{r}
# plot silhouette width
# pdf(file.path(figure_dir, "pam_all_data.pdf"))
plotSilhouette(all_gower_mat, k = 2, n = 15)
# dev.off()

# clustering with PAM
all_pam_dat <- pam(all_gower_dist, diss = TRUE, k = 2)
all_pam_clusters <- factor(all_pam_dat$clustering)

# plot tnse
# pdf(file.path(figure_dir, "tsne_all_data.pdf"), width = 16, height = 8)
age_level <- factor(
  ifelse(metadata$challengeAge > (2022-2016), "old", "young"), 
  levels = c("old", "young"))
plotDistTSNE(all_gower_dist, 
             cluster = all_pam_clusters, 
             label = age_level,
             color = age_level,
             color.legend.title = "Age Level",
             seed = 1234) 
# dev.off()
```

Since user engagement data was not well applied until 2016, it makes sense that there are two clear clusters of challenges by the ages. 

#### Feature summary by clusters
```{r, include=FALSE}
all_quan_dat <- metadata[, num_inx]
# all_quan_dat <- sapply(all_quan_dat, as.numeric)
# all_quan_dat <- as.data.frame(log2(all_quan_dat + 0.1))
all_quan_dat$clusters <- as.factor(all_pam_clusters)
```

#### Boxplot by clusters
```{r,fig.height=8}
# quantitative features
output_feat <-  c("numberTotalSubmissions", "numberOfTotalRegistrants")
ue_plot_feat <- c("numberOfAuthors", "numberOfFollowers", "numberOfThreads", 
                  "maxRepliesPerThread", "maxViewsPerThread", "AvgResponsesPerCommenter")
feature_to_plot <- c("challengeAge", output_feat, ue_plot_feat)
bp_all_data <- plotQuanBoxplot(select(all_quan_dat, c("clusters",all_of(feature_to_plot))), 
                               test.method = "wilcox.test", pos.x = 0.18, pos.y = 0.98
                               )
bp_all_data
if (update_res) ggsave(file.path(figure_dir, "boxplot_all_data.pdf"), bp_all_data, width = 10, height = 8)
```

```{r}
# qualitative features
qual_data_all <- lapply(c(cate_feats, logi_feats), function(col) {
 data.frame(metadata[, col]) %>%
    replace_na("Unknown") %>%
    set_names(col) %>%
    splitstackshape::cSplit_e(col, sep= ',', type = 'character', fill = 0, drop = TRUE)
}) %>% bind_cols()
# count number of values per cluster
qual_data_all$clusters <- as.factor(all_pam_clusters)
qual_counts_all <- qual_data_all %>% 
  gather(key, value, -clusters) %>% 
  group_by(clusters, key, .drop = TRUE) %>% 
  summarise(n = sum(value)) %>%
  spread(key, n) %>%
  tibble::column_to_rownames("clusters")
# calculate fisher exact test for each category
qual_res_all <-sapply(c(cate_feats, logi_feats), function(col) {
  contingency <- qual_counts_all %>% select(starts_with(col))
  res <- fisher.test(contingency)
  res$p.value
}) %>% 
  set_names(c(cate_feats, logi_feats)) %>%
  p.adjust("BH")
qual_res_all
#         dataTypeTiers          sponsorTiers      contributorTiers 
#            0.49474496            0.42975553            0.02753499 
# modelContainerization     monetaryIncentive     speakingIncentive 
#            0.02753499            1.00000000            0.27029760 

# frequency plot
cols_fun <- circlize::colorRamp2(c(0, 10), c("white", "#1b3f61"))
hp_frq_dat <- select(qual_counts_all,-c("speakingIncentive_FALSE", "monetaryIncentive_FALSE", "modelContainerization_FALSE")) %>% select(-ends_with("Unknown"))
tmp <- hp_frq_dat
inx <- which(grepl("FALSE|TRUE", colnames(hp_frq_dat)))
colnames(tmp)[-inx] <- strsplit(colnames(hp_frq_dat)[-inx], "_") %>% sapply(., `[[`, 2)
colnames(tmp)[inx] <- strsplit(colnames(hp_frq_dat)[inx], "_") %>% sapply(., `[[`, 1)
hp_frq_dat <- tmp
hp_freq_all <- pheatmap(t(hp_frq_dat), cluster_rows = F, cluster_cols = F, color = cols_fun(1:10), angle_col = 0, fontsize = 12)
if (update_res) ggsave(file.path(figure_dir, "heatmap_frequency_all.pdf"), hp_freq_all, width = 3, height = 6)
```

### Young challenges Only

#### Analysis on the challenges from cluster 2 (young challenges)

```{r}
new_chlg_dat <- all_dat[all_pam_clusters == 2, ]
# remove age as well
new_chlg_dat <- new_chlg_dat[,-1]

# re-calculate gower distance with subsetted data
new_chlg_gower_dist <- 
  daisy(new_chlg_dat,
        metric = "gower" ,
        type = list(logratio = 2))
new_chlg_gower_mat <- as.matrix(new_chlg_gower_dist)
```

#### Sub-clustering
```{r}
# pdf(file.path(figure_dir, "pam_remove_old.pdf"))
plotSilhouette(new_chlg_gower_mat, k = 5, n = 26)
# dev.off()
```

```{r}
new_chlg_pam_dat <- pam(new_chlg_gower_dist, diss = TRUE, k = 5)
new_chlg_pam_clusters <- factor(new_chlg_pam_dat$clustering)
```

#### TSNE
```{r}
# pdf(file.path(figure_dir, "tsne_remove_old.pdf"), width = 7, height = 5)
new_years <- 2022 - as.numeric(all_dat[all_pam_clusters == 2, "challengeAge"])
plotDistTSNE(new_chlg_gower_dist, 
             cluster = new_chlg_pam_clusters, 
             label = new_years,
             color.legend.title = "clusters",
             seed = 1)
# dev.off()
```

```{r}
new_chlg_quan_dat <- metadata[all_pam_clusters == 2, num_feats]
new_chlg_quan_dat$clusters <- as.factor(new_chlg_pam_clusters)
```

#### Feature boxplot by clusters
```{r}
# quantitative features
bp_remove_old <- plotQuanBoxplot(new_chlg_quan_dat[, c("clusters", feature_to_plot)], test.method = "anova")
bp_remove_old
if (update_res) ggsave(file.path(figure_dir, "boxplot_remove_old.pdf"), bp_remove_old, width = 10, height = 7)
```


```{r}
# qualitative features
qual_data_rm_old <- lapply(c(cate_feats, logi_feats), function(col) {
 data.frame(metadata[all_pam_clusters == 2, col]) %>%
    replace_na("Unknown") %>%
    set_names(col) %>%
    splitstackshape::cSplit_e(col, sep= ',', type = 'character', fill = 0, drop = TRUE)
}) %>% bind_cols()
# count number of values per cluster

qual_data_rm_old$clusters <- as.factor(new_chlg_pam_clusters)
# qual_data_rm_old <- qual_data_rm_old[, -grep("Unknown", colnames(qual_data_rm_old))]
qual_counts_rm_old <- qual_data_rm_old %>%
  gather(key, value, -clusters) %>% 
  group_by(clusters, key, .drop = TRUE) %>% 
  summarise(n = sum(value)) %>%
  spread(key, n) %>%
  tibble::column_to_rownames("clusters")

# calculate fisher exact test for each category
qual_res_rm_old <-sapply(c(cate_feats, logi_feats), function(col) {
  contingency <- qual_counts_rm_old %>% select(starts_with(col))
  res <- fisher.test(contingency, simulate.p.value = T)
  res$p.value
}) %>% 
  set_names(c(cate_feats, logi_feats)) %>%
  p.adjust("BH")
qual_res_rm_old
    #     dataTypeTiers          sponsorTiers      contributorTiers modelContainerization 
    #       0.504947526           0.504947526           0.990504748           0.005997001 
    # monetaryIncentive     speakingIncentive 
    #       0.504947526           0.124437781 
# frequency plot
cols_fun <- circlize::colorRamp2(c(0, 10), c("white", "#1b3f61"))
hp_frq_dat <- select(qual_counts_rm_old,-c("speakingIncentive_FALSE", "monetaryIncentive_FALSE", "modelContainerization_FALSE")) %>% select(-ends_with("Unknown"))
hp_freq_rm_old_rm_ue <- pheatmap(t(hp_frq_dat), cluster_rows = F, cluster_cols = F, color = cols_fun(1:10), angle_col = 0)

if (update_res) ggsave(file.path(figure_dir, "heatmap_frequency_remove_old_rm_ue.pdf"), hp_freq_rm_old_rm_ue, width = 4, height = 6)
```
